Contributions of adaptation and purifying selection to SARS-CoV-2 evolution

### 2022-08-18 manuscript/variant_rates.tex
comments by louis and Jesse

\begin{abstract}
Continued evolution and adaptation of SARS-CoV-2 has lead to more transmissible and immune-evasive variants with profound impact on the course of the pandemic.
Here we analyze the evolution of the virus over 2.5 years since its emergence and estimate rates of evolution for synonymous and non-synonymous changes.
We estimate separate rates for evolution within clades, that is well defined mono-phyletic groups with gradual evolution, and for the pandemic overall.
The rate of synonymous mutations is found to be around 6 changes per year.
Rates within variants vary little and are compatible with a overall rate.
In contrast, the rate at which variants accumulate amino acid changes was initially around 12-16 changes per year, but in 2021 and 2022 dropped to 6-9 changes per year.
The overall rate of non-synonymous evolution, that is across variants, is estimated to be about 25 amino acid changes per year.
This 2-fold higher rate indicates that adaptive evolutionary process that gave rise to the different variants is qualitatively different from that in typical transmission chains.
We further quantify the spectrum of mutations and purifying selection in different SARS-CoV proteins.
Many accessory proteins evolve under limited evolutionary constraint with little short term purifying selection.
About half of mutations in other proteins are strongly deleterious and rarely observed, even at low frequency.
\end{abstract}

@@

Since its emergence in late 2019 \citep{zhu_novel_2020}, SARS-CoV-2 has displayed a discontinuous pattern of evolution with large jumps in sequence space giving rise to phylogenetically distinct variants \citep{hodcroft_spread_2021,volz_assessing_2021,tegally_detection_2021,faria_genomics_2021,naveca_covid-19_2021,viana_rapid_2022}.
Many of these variants spread considerably faster and quickly displaced the resident variants either because of intrinsically increased transmissibility, evasion of prior immunity in the population, or a combination of both.
Specific Variants of Concern (VoC) or Interest (VoI) were designated by the WHO and labeled by Greek letters \citep{konings_sars-cov-2_2021}.
The branches leading to these variants are characterized by many amino acid changing mutations that often cluster in the S1 domain of the spike protein \citep{kistler_rapid_2022}.

This pattern of rapid non-synonymous evolution in viral surface proteins that interact with the host cells is common among many RNA viruses and for example well studied in influenza virus A evolution \citep{bhatt_genomic_2011,strelkowa_clonal_2012}.
But adaptive evolution of influenza viruses tends to be gradual without large jumps in sequence space, while new variants of SARS-CoV-2 with 10s of novel mutations emerged suddenly without intermediate genomes being observed, the most dramatic being the emergence of Omicron in late 2021 \citep{viana_rapid_2022}.
The pre-dominant hypothesis for this cryptic emergence of highly mutated, transmissible and immune evasive variants are chronic infections that are common in patients with impaired immune systems, either through HIV-1 infection \citep{cele_sars-cov-2_2022} or medical intervention \citep{choi_persistence_2020,kemp_sars-cov-2_2021}.
Onward transmission from such chronic infections has also been documented \citep{gonzalez-reiche_intrahost_2022}.
However, to date there is no direct evidence for the mode of emergence of any variant.
The case for chronic infection being an important contributor is strongest for the variants Alpha and Omicron \citep{hill_origins_2022}.

The saltatory nature of SARS-CoV-2 evolution implies that evolution within individual variants, clades, or lineages might have different characteristics from the global pattern, which is dominated by atypical bursts of evolution leading to new variants.
\citet{tay_emergence_2022} and \citet{hill_origins_2022} show that the rate of evolution along these inter-clade branches is up to four-fold higher than the background rate.
Here, we build on these results and investigate the patterns of SARS-CoV-2 diversification within variants and compare these to the global dynamics of evolution and adaptation.
This comparison reveals a consistent dichotomy between slow within-variant evolution and rapid adaptive evolution giving rise to new variants.
This different in evolutionary rate is only seen for non-synonymous changes -- the rate of synonymous evolution within variants is compatible with that seen between variants.
Furthermore, early variants display more rapid non-synonymous evolution than later variants suggesting more ubiquitous adaptive evolution early on.
We further quantify the level of functional constraint of different open reading frames and infer a map of mutational tolerance across the genome.

\begin{figure*}[tb]
    \includegraphics[width=\textwidth]{figures/rtt/20C_rtt.pdf}
    \includegraphics[width=\textwidth]{figures/counts/20C_counts.pdf}
    \caption{{\bf Within-clade divergence increases linearly with time.} Top: Each panel shows the number of within-clade mutations (total (A), amino acid changing (B), synonymous (C)) as a function of time.
    The green line in panel A indicates the divergence cut-off, panels B\&C only show sequences that pass the divergence filter. Each panel also shows mean $\pm$ standard deviation and a weighted linear fit. Analogous figures for all clades considered are included in the appendix.
    Bottom: Panels D and E show the prevalence of specific genotypes (D) and specific mutations (E). In the case shown, the founder genotype initially dominates and no daughter genotype or mutation dominate. Panel F shows a Poisson model fit to the breakdown of the population into genotypes with different number of mutations over time.
    \label{fig:within_clade}}
\end{figure*}

@@

Evolutionary rates and divergence times are typically estimated using phylogenetic \citep{drummond_relaxed_2006}.
These methods, however, can not handle the volume of SARS-CoV-2 data available and data have to be dramatically down-sampled.
Furthermore, phylogenetic methods impose an hierarchical structure on the data and are thus very sensitive to problematic sequences or metadata: Any mis-dated, recombinant, contaminated, or otherwise chimeric sequence can fatally distort the analysis, in particular if placed early in tree.

To circumvent many of the above mentioned problems and still use the majority of the available data, we use a combination of automated filtering and simple robust approaches to analyze the evolutionary patterns (see Materials and Methods).
We first use Nextclade \citep{aksamentov_nextclade_2021} to assign all sequences to one of the Nextstrain clades \citep{hadfield_nextstrain_2018}, which are denoted by a Year-Letter combination (e.g.~19A, 20A, 20B, \ldots).
Sequences belonging to recombinant Pango lineages are excluded \citep{rambaut_dynamic_2020}.
These clades represent well defined groups of sequences with little non-vertical evolution within them and we analyze data from each clade separately.
Furthermore, we define a ``founder'' genotype for each clade and exclude any sequence that does not have the full set of mutations relative to the reference sequence Wuhan/Hu-1.
This founder sequence is manually curated for each clade considered.
This filtering removes most incomplete sequences as well as sequences where amplicon drop-out are back-filled with reference sequence.
For this reduced set of sequences, we determine the mutations they carry on top of the founder genotype to analyze diversification and divergence within the clade.
The latter step is done for nucleotide changes as well as for amino acid changes.

Within each clade, the number of mutations is expected to increase linearly in time and the variation around this mean would, in an ideal case, obey Poisson statistics.
For the majority of sequences, this is approximately true, but some problematic sequences have more mutations than expected.
To exclude these outliers, we perform a simple linear regression of the number of ``intra-clade'' mutations against time and remove sequences whose deviation from the linear fit exceeds twice the expected standard deviation by 3.

After removing these outliers, we now bin the data by calendar week and determine the mean and standard deviation in each bin.
Evolutionary rate and putative emergence date of the variant are then estimated by weighted linear regression where each time bin is weighted with the fourth root of the number of sequences in the bins.
The exact functional form of this weighing does not have a big influence on the results, but a sublinear weighing helps to counter the large variation in sequencing effort across countries and the unavoidable imbalance due to the fact that few sequences are available early.
Fig.~\ref{fig:within_clade} shows the increasing intra-clade divergence for clade 20C, corresponding to a large clade that emerged in early 2020 that was common in North America and Scandinavia.
Both synonymous and non-synonymous within-clade average divergence increase linearly over time allowing for a robust estimate of the rate.

Due to shared ancestry, divergences of sequences are not independent data points and a regression against time is not suited to estimate evolutionary rates in general.
In particular, confidence intervals are difficult to obtain.
However, in the case of rapidly expanding variants we typically observe a large number of independent lineages emanating from one or several basal polytomies.
Along each of these lineages, mutation accumulation is independent.
While not every sequence is an independent sample, the effective number of independent sampled is large and the steadily increasing average divergence allows to estimate the rate robustly.

A simple model for diversity within a growing variant is a super-critical branching process with growth rate $\alpha$ and an embedded mutation process.
Offspring of genomes with $i$ mutations contain $i+j$ mutations, where $j$ is a Poisson distributed number with mean $\mu t$ (mutation rate $\mu$ and generation time $t$).
The probability that offspring genomes are identical to their parents is $u = 1-e^{-\mu t}$, which for a generation time of $t=5days$ and a rate of $\mu = 15/year$ evaluates to $u\approx 0.2$.

When considering a rapidly growing well sampled outbreak, typically a single founder genotype will give rise to a large number of daughter lineages that evolve independently.
In this case, the diversification processes is robustly described by its mean.
Since the above branching process is linear, the mean number of cases $n$ will increase exponentially with rate $\alpha$, while the number of genomes with $i$ mutations relative to the founder $m_i$ grows with rate $\alpha - u$ per generation:
\begin{equation}
    \begin{split}
        \frac{dn}{dt} &= \alpha n \\
        \frac{dm_0}{dt} &= (\alpha - u) m_0 \\
        \frac{dm_1}{dt} &= u m_0 (\alpha - u) m_1 \\
        \frac{dm_2}{dt} &= u m_1 (\alpha - u) m_2 \\
        \cdots
    \end{split}
\end{equation}
with solution $m_i = e^{(\alpha - u)t} \frac{(ut)^i}{i!}$.
At each time the number of mutations in the population is expected to Poisson distributed with mean $ut$, while the overall number of cases is $e^{\alpha t}$ or more generally $e^{\int \alpha(t) dt}$ if growth rate or ascertainment varies over time.

For some clades, especially those that are well sampled soon after their emergence, this Poisson model is good fit to diversity accumulation and yields estimates of rates and time of origin that are compatible with the divergence regression, see Fig.~\ref{fig:within_clade}F for clade 20C.
In this case, the founder genotype initially dominates, but is gradually replaced -- first by single mutant genotypes, then double mutants, and so forth (comp.~Fig.~\ref{fig:within_clade}).
Analogous graphs for all other clades considered are included in the appendix.

In some cases, notable subclades do become dominant in the early stochastic dynamic.
This can result in increased within clade divergence, effectively generating an offset to diversity.
Extrapolation of the linear fit to zero divergence is thus not necessarily a good estimator of the emergence time of a variant.
For several clades 19B, 20H (Beta), 21D (Eta), 21G (Lambda), 21 (Mu), 21I (Delta), 21J (Delta), 21L (Omicron, BA.2), and 22B (Omicron, BA.5), founder like variants are a minority even in early data.

These Poisson weights of mutation numbers are again only valid if mutations accumulate along many independent lineages.
In particular, this assumption is violated if some lineages spread systematically faster than others either because of epidemiological factors or the mutations they carry.
In variant 21K (Omicron, BA.1), a sublineage with mutation \texttt{S:R346K} might have enjoyed a transmission advantage.
A possible consequence of this advantage is visible in Supp.~Fig.~\ref{fig:21K_divergence} where genotypes with 0, 1, or 2 mutations are decaying more rapidly than those with more mutations.

\begin{figure*}
    \includegraphics[width=0.8\textwidth]{figures/rate_summary.pdf}
    \caption[]{{\bf Divergence and evolutionary rates of different Nextstrain clades.} Panels A,B,\&C show the estimated divergence of the founder genotype of each clade (big dot) and the subsequent divergence trend for all nucleotide changes, amino acid changes, and synonymous changes, respectively. In addition, each panel contains a regression of the divergence of clade founders vs time (gray line).
    The standard deviation expected based on Poisson statistics is indicated as shaded area.
    Panel D summarizes the individual rate estimates (dots and squares) and compares them to the estimate inter-clade rates (gray lines).
    \label{fig:rate_summary} }
\end{figure*}

Despite these caveats, for almost all major Nextstrain clades, slope at which diversity increases is robust (similar patterns as Fig.~\ref{fig:within_clade}A-C for clade 20C), allowing us to estimate clade specific evolutionary rates for amino acid and synonymous changes.
These rates are summarized in Fig.~\ref{fig:rate_summary} and Tab.~\ref{tab:rates}.
In addition, we estimate the rate at which the variants themselves accumulated amino acid and synonymous changes by regressing the number of changes of the variants (relative to the putative root in clade 19B \citep{caraballo-ortiz_tophap_2022}) against the estimated time of origin of the variants.
These regressions are shown in thick gray lines in Fig.~\ref{fig:rate_summary}.

Rates of synonymous change are very consistent across variants (about 5-8 changes per genome per year) and also agree with the overall rate of synonymous changes of 7.3 changes per genome per year.
The rates of non-synonymous changes are much more variable (Fig.~\ref{fig:rate_summary}D).
Within clades, the rate of non-synonymous changes varies between 5 and 16 changes per year.
Earlier clades are estimated to have larger rates around 10-15 changes per year, while rate estimates for later clade fall  between 3 to 9 changes per year Fig.~\ref{fig:rate_progression}.
In contrast, the inter-clade non-synonymous rate exceeds 25 changes per year.

Nextstrain clades tend to be defined by long branches leading to a large polytomy.
It could thus be that the estimated inter-clade rate exceeds the intra-clade rate purely because of this conditioning.
This effect might be particularly important early on in the pandemic where branches with as little two mutations were used to define new clades.
We therefore also included composite clades 19B+, 19B++ and 20A+ containing sequences from 19A and 19B (rooted on 19B), 19A, 19B, 20A, 20B, 20C, 20D (rooted on 19B), and 20A, 20B, 20C, 20D (rooted on 20A).
The estimates for composite clades 19B+ and 20A+ are consistent with the estimates of the individual clades, while apparent rates of 19B++ are considerably higher.
The latter is due to the rapid expansion and subsequent dominance of clade 20A and its descendants which rapidly added four additional mutations \citep{korber_tracking_2020}.
We argue that this is an early example of an accelerated global rate of evolution due to adaptive evolution, see Discussion.

At the other extreme, some Omicron clades have very low rate estimates which should be interpreted carefully.
Founder genotypes of Omicron clades 21L (BA.2), 22A (BA.4) and 22B (BA.5) are barely sampled.
These clades showed considerable diversity, including reversions to the reference, soon after their initial discovery \citep{tegally_emergence_2022}.
This diversity and the complex, possibly recombinant, origin of the Omicron clades makes estimating their rate of evolution challenging.

\begin{table*}
\begin{tabular}{l|rrrrrr}
    \hline
    {\bf clade} &  overall rate $[y^{-1}]$ & aa rate $[y^{-1}]$ &  syn rate $[y^{-1}]$ & overall div. &  aa div. &  syn div. \\
    \hline
    19B &     19.46 &    12.37 &      7.09 &        0 &       0 &        0 \\
   19B+ &     18.56 &    11.92 &      6.64 &        0 &       0 &        0 \\
  19B++ &     27.91 &    16.41 &     11.50 &        0 &       0 &        0 \\
  19A &     19.73 &    12.79 &      6.95 &        2 &       1 &        1 \\
  20A &     15.08 &     7.94 &      7.15 &        6 &       3 &        3 \\
    20B &     14.29 &     7.69 &      6.59 &        9 &       5 &        4 \\
    20C &     16.73 &     9.92 &      6.81 &        8 &       5 &        3 \\
   20A+ &     16.30 &     9.40 &      6.90 &        6 &       3 &        3 \\
    20E &     14.04 &     5.65 &      8.39 &       13 &       5 &        8 \\
    20H &     15.48 &     9.51 &      5.96 &       21 &      16 &        5 \\
    20I &     11.56 &     5.49 &      6.07 &       30 &      19 &       11 \\
    20J &     12.97 &     6.29 &      6.68 &       35 &      22 &       13 \\
    21D &     11.36 &     6.29 &      5.07 &       27 &      13 &       14 \\
    21G &     12.04 &     6.51 &      5.53 &       28 &      19 &        9 \\
    21H &     18.25 &    10.05 &      8.20 &       31 &      21 &       10 \\
    21I &     16.55 &     8.93 &      7.62 &       30 &      24 &        6 \\
    21J &     14.30 &     7.17 &      7.14 &       34 &      28 &        6 \\
    21K &     13.48 &     6.45 &      7.03 &       56 &      46 &       10 \\
    21L &      9.24 &     4.39 &      4.85 &       68 &      50 &       18 \\
    22A &      7.71 &     2.81 &      4.90 &       71 &      52 &       19 \\
    22B &     11.90 &     5.03 &      6.87 &       66 &      50 &       16 \\
           \hline
    \end{tabular}
\caption{{\bf Evolutionary rates estimates from root-to-tip regressions for overall nucleotide changes, amino acid changes, and synonymous changes.}
    The last three columns give the distances of the clade founder sequence from putative MRCA of SARS-CoV-2.
\label{tab:rates}}
\end{table*}

\begin{figure}
    \includegraphics[width=0.5\textwidth]{figures/rate_progression.pdf}
    \caption{{\bf Divergence and evolutionary rates of different Nextstrain clades over time.} Synonymous rates estimates are stable in time and fluctuate around the rate estimates for between clades. Non-synonymous rate estimates are highest for clades 19A - 20C.
    \label{fig:rate_progression} }
\end{figure}

@@

 \begin{figure*}[tb]
    \includegraphics[width=\textwidth]{figures/fitness_landscape.pdf}
    \caption{{\bf Landscape of selective constraint along the SARS-CoV-2 genome.}
    Solid lines show sliding window smoothing of estimated mutational tolerance at 1st, 2nd, and 3rd positions for a window size of 20 and 7 (faint lines) sites.
    The markers below the lines are a qualitative indication of variable positions across SARS-CoV-2 diversity.
    \label{fig:fitness_landscape}}
 \end{figure*}

>>>>

\begin{abstract}
Continued evolution and adaptation of SARS-CoV-2 has lead to more transmissible and immune-evasive variants with profound impact on the course of the pandemic.
Here we analyze the evolution of the virus over 2.5 years since its emergence and estimate rates of evolution for synonymous and non-synonymous changes.
We estimate separate rates for evolution within clades, that is well defined mono-phyletic groups with gradual evolution, and for the pandemic overall.
The rate of synonymous mutations is found to be around 6 changes per year.
Rates within variants vary little and are compatible with the overall rate.
In contrast, the rate at which variants accumulate amino acid changes (non-synonymous mutation) was initially around 12-16 changes per year, but in 2021 and 2022 dropped to 6-9 changes per year.
The overall rate of non-synonymous evolution, that is across variants, is estimated to be about 25 amino acid changes per year.
This 2-fold higher rate indicates that adaptive evolutionary process that gave rise to the different variants is qualitatively different from that in typical transmission chains.
We further quantify the spectrum of mutations and purifying selection in different SARS-CoV proteins.
Many accessory proteins evolve under limited evolutionary constraint with little short term purifying selection.
About half of the mutations in other proteins are strongly deleterious and rarely observed, even at low frequency.
\end{abstract}

@@

Since its emergence in late 2019 \citep{zhu_novel_2020}, SARS-CoV-2 has displayed a discontinuous pattern of evolution with large jumps in sequence space giving rise to phylogenetically distinct variants \citep{hodcroft_spread_2021,volz_assessing_2021,tegally_detection_2021,faria_genomics_2021,naveca_covid-19_2021,viana_rapid_2022}.
Many of these variants spread considerably faster and quickly displaced the resident variants either because of intrinsically increased transmissibility, evasion of prior immunity in the population, or a combination of both.
Specific Variants of Concern (VoC) or Interest (VoI) were designated by the WHO and labeled by Greek letters \citep{konings_sars-cov-2_2021}.
The branches leading to these variants are characterized by many amino acid-changing mutations that often cluster in the S1 domain of the spike protein \citep{kistler_rapid_2022}.

This pattern of rapid non-synonymous evolution in viral surface proteins that interact with the host cells is common among many RNA viruses and for example well studied in influenza A virus evolution \citep{bhatt_genomic_2011,strelkowa_clonal_2012}.
But adaptive evolution of influenza viruses tends to be gradual without large jumps in sequence space, while new variants of SARS-CoV-2 with tens of novel mutations emerged suddenly without intermediate genomes being observed, the most dramatic being the emergence of Omicron in late 2021 \citep{viana_rapid_2022}.
The pre-dominant hypothesis for this cryptic emergence of highly mutated, transmissible and immune evasive variants are chronic infections that are common in patients with impaired immune systems, either through HIV-1 infection \citep{cele_sars-cov-2_2022} or medical intervention \citep{choi_persistence_2020,kemp_sars-cov-2_2021}.
Onward transmission from such chronic infections has also been documented \citep{gonzalez-reiche_intrahost_2022}.
However, to date there is no direct evidence for the mode of emergence of any variant.
The case for chronic infection being an important contributor is strongest for the variants Alpha and Omicron \citep{hill_origins_2022}.

The dichotomous pattern of SARS-CoV-2 evolution with step-wise evolution within clades or variants and atypical bursts of evolution leading to new variants has been investigated by \citet{tay_emergence_2022} and \citet{hill_origins_2022}, who show that the rate of evolution along branches giving rise to new variants is up to four-fold higher than the background rate.
Here, we build on these results and investigate the patterns of SARS-CoV-2 diversification within variants and compare these to the global dynamics of evolution and adaptation.
This comparison reveals a consistent dichotomy between slow within-variant evolution and rapid adaptive evolution giving rise to new variants.
This difference in evolutionary rate is only seen for non-synonymous changes -- the rate of synonymous evolution within variants is compatible with that seen between variants.
Furthermore, early variants display more rapid non-synonymous evolution than later variants suggesting more ubiquitous adaptive evolution early on.
We further quantify the level of functional constraint of different open reading frames and infer a map of mutational tolerance across the genome.

\begin{figure*}[tb]
    \includegraphics[width=\textwidth]{figures/rtt/20C_rtt.pdf}
    \includegraphics[width=\textwidth]{figures/counts/20C_counts.pdf}
    \caption{{\bf Within-clade divergence increases linearly with time (example clade 20C).} Top: Each panel shows the number of within-clade mutations (total (A), amino acid changing (B), synonymous (C)) as a function of time.
    The green line in panel A indicates the divergence cut-off, panels B\&C only show sequences that pass the divergence filter. Each panel also shows mean $\pm$ standard deviation and a weighted linear fit.
    Bottom: Panels D and E show the prevalence of the most common genotypes (D) and mutations (E) during the first 3 months. In the case shown (20C), the founder genotype initially dominates and no specific mutation ever reaches high frequency. Note that specific genotypes invariably decrease in frequency due to subsequent mutations, while mutations irrespective of their genetic background persist. Panel F shows a Poisson model fit to the breakdown of the population into genotypes with different number of mutations over time. Analogous figures for other clades are included in the appendix.
    \label{fig:within_clade}}
\end{figure*}

@@

Evolutionary rates and divergence times are typically estimated using phylogenetic approaches \citep{drummond_relaxed_2006}.
These methods, however, cannot handle the volume of SARS-CoV-2 data available and data have to be dramatically down-sampled.
Furthermore, phylogenetic methods impose an hierarchical structure on the data and are thus very sensitive to problematic sequences or metadata: Any mis-dated, recombinant, contaminated, or otherwise chimeric sequence can fatally distort the analysis, in particular if placed early in tree.
Problematic sequences can be particularly common when a new variant takes over since sequencing protocols need adjusting \citep{deMaio_issues_2020}.

To circumvent many of the above-mentioned problems and still use the majority of the available data, we use a combination of automated filtering and simple robust approaches to analyze the evolutionary patterns (see Materials and Methods).
We first use Nextclade \citep{aksamentov_nextclade_2021} to assign all sequences to one of the Nextstrain clades \citep{hadfield_nextstrain_2018}, which are denoted by a Year-Letter combination (e.g.~19A, 20A, 20B, \ldots).
Sequences belonging to recombinant Pango lineages are excluded \citep{rambaut_dynamic_2020}.
These clades represent well-defined groups of sequences with little evidence of recombination within them and we analyze data from each clade separately.
Furthermore, we define a ``founder'' genotype for each clade and exclude any sequence that does not have the full set of mutations relative to the reference sequence Wuhan/Hu-1.
This founder sequence is manually curated for each clade considered.
This filtering removes most incomplete sequences as well as sequences where amplicon drop-out are back-filled with the reference sequence.
For this reduced set of sequences, we determine the mutations they carry on top of the founder genotype to analyze diversification and divergence within the clade.
The latter step is done for nucleotide changes as well as for amino acid changes.

Within each clade, the number of mutations is expected to increase linearly in time and the variation around this mean would, in an ideal case, obey Poisson statistics.
For the majority of sequences, this is approximately true, but some problematic sequences have more mutations than expected.
To exclude these outliers, we perform a simple linear regression of the number of ``intra-clade'' mutations against time and remove sequences whose deviation from the linear fit exceeds twice the expected standard deviation by 3 mutations.

After removing these outliers, we now bin the data by calendar week and determine the mean and standard deviation in each bin.
Evolutionary rate and putative emergence date of the variant are then estimated by weighted linear regression where each time bin is weighted with the fourth root of the number of sequences in the bins.
The exact functional form of this weighting does not have a big influence on the results, but a sublinear weighting helps to counter the large variation in sequencing effort across countries and the unavoidable imbalance due to the fact that few sequences are available early after a new variant emerges.
Fig.~\ref{fig:within_clade} shows the increasing intra-clade divergence for clade 20C, corresponding to a large clade that emerged in early 2020 that was common in North America and Scandinavia.
Both synonymous and non-synonymous within-clade average divergence increase linearly over time allowing for a robust estimate of the rate.

Due to shared ancestry, divergences of sequences are not independent data points and a regression against time is not suited to estimate evolutionary rates in general.
In particular, confidence intervals are difficult to obtain.
However, in the case of rapidly expanding variants we typically observe a large number of independent lineages emanating from one or several basal polytomies.
Along each of these lineages, mutation accumulation is independent.
While not every sequence is an independent sample, the effective number of independent samples is large and the steadily increasing average divergence allows to estimate the rate robustly.

A simple model for diversity within a growing variant is a super-critical branching process with growth rate $\alpha$ and an embedded mutation process.
Offspring of genomes with $i$ mutations contain $i+j$ mutations, where $j$ is a Poisson distributed number with mean $\mu t$ (mutation rate $\mu$ and generation time $t$).
The probability that offspring genomes are identical to their parents is $u = 1-e^{-\mu t}$, which for a generation time of $t=5$ days and a rate of $\mu = 15/year$ evaluates to $u\approx 0.2$ (note that this rate excludes strongly deleterious mutations, see Discussion).

When considering a rapidly growing well sampled outbreak, typically a single founder genotype will give rise to a large number of daughter lineages that evolve independently.
In this case, the diversification processes is robustly described by its mean.
Since the above branching process is linear, the mean number of cases $n$ will increase exponentially with rate $\alpha$, while the number of genomes with $i$ mutations relative to the founder $m_i$ grows with rate $\alpha - u$ per generation:
\begin{equation}
    \begin{split}
        \frac{dn}{dt} &= \alpha n \\
        \frac{dm_0}{dt} &= (\alpha - u) m_0 \\
        \frac{dm_1}{dt} &= u m_0 (\alpha - u) m_1 \\
        \frac{dm_2}{dt} &= u m_1 (\alpha - u) m_2 \\
        \cdots
    \end{split}
\end{equation}
with solution $m_i = e^{(\alpha - u)t} \frac{(ut)^i}{i!}$.
At each time the number of mutations in the population is expected to be Poisson distributed with mean $ut$, while the overall number of cases is $e^{\alpha t}$ or more generally $e^{\int \alpha(t) dt}$ if growth rate or ascertainment varies over time.

For some clades, especially those that are well sampled soon after their emergence, this Poisson model is a good fit to diversity accumulation and yields estimates of rates and time of origin that are compatible with the divergence regression, see Fig.~\ref{fig:within_clade}F for clade 20C.
In this case, the founder genotype initially dominates, but is gradually replaced -- first by single mutant genotypes, then double mutants, and so forth (comp.~Fig.~\ref{fig:within_clade}).
Analogous graphs for all other clades considered are included in the appendix.

In other cases, notable subclades do become dominant in the early stochastic dynamic.
This can result in increased within-clade divergence, effectively generating an offset to diversity.
Extrapolation of the linear fit to zero divergence is thus not necessarily a good estimator of the emergence time of a variant.
For several clades 19B, 20H (Beta), 21D (Eta), 21G (Lambda), 21 (Mu), 21I (Delta), 21J (Delta), 21L (Omicron, BA.2), and 22B (Omicron, BA.5), founder-like variants are a minority even in early data.

These Poisson weights of mutation numbers are again only valid if mutations accumulate along many independent lineages.
In particular, this assumption is violated if some lineages spread systematically faster than others either because of epidemiological factors or because they carry adaptive mutations.
In variant 21K (Omicron, BA.1), a sublineage with mutation \texttt{S:R346K} might have enjoyed a transmission advantage.
A possible consequence of this advantage is visible in Supp.~Fig.~\ref{fig:21K_divergence} where genotypes with 0, 1, or 2 mutations are decaying more rapidly than those with more mutations.

\begin{figure*}
    \includegraphics[width=0.8\textwidth]{figures/rate_summary.pdf}
    \caption[]{{\bf Divergence and evolutionary rates of different Nextstrain clades.} Panels A,B,\&C show the estimated divergence of the founder genotype of each clade (big dot or square) and the subsequent divergence trend for all nucleotide changes, amino acid changes, and synonymous changes, respectively.
    In addition, each panel contains a regression of the divergence of clade founders vs time (gray line).
    The standard deviation expected based on Poisson statistics is indicated as shaded area.
    Panel D summarizes the individual rate estimates (dots and squares) and compares them to the estimate ofinter-clade rates (gray lines).
    In panel D, clades are in alphabetical order, which is similar to their (uncertain) order of emergence.
    \label{fig:rate_summary} }
\end{figure*}

Despite these caveats, for almost all major Nextstrain clades, the slope at which diversity increases is robust (similar patterns as Fig.~\ref{fig:within_clade}A-C for clade 20C), allowing us to estimate clade-specific evolutionary rates for amino acid and synonymous changes.
These rates are summarized in Fig.~\ref{fig:rate_summary} and Tab.~\ref{tab:rates}.
In addition, we estimate the rate at which the variants themselves accumulated amino acid and synonymous changes by regressing the number of changes of the variants (relative to the putative root in clade 19B \citep{caraballo-ortiz_tophap_2022}) against the estimated time of origin of the variants.
These regressions are shown in thick gray lines in Fig.~\ref{fig:rate_summary}.

Rates of synonymous change are very consistent across variants (about 5-8 changes per genome per year) and also agree with the overall rate of synonymous changes of 7.3 changes per genome per year.
The rates of non-synonymous changes are much more variable (Fig.~\ref{fig:rate_summary}D).
Within clades, the rate of non-synonymous changes varies between 5 and 16 changes per year.
Earlier clades are estimated to have larger rates around 10-15 changes per year, while rate estimates for later clade fall  between 3 to 9 changes per year (see Fig.~\ref{fig:rate_progression}).
In contrast, the inter-clade non-synonymous rate exceeds 25 changes per year.

Nextstrain clades tend to be defined by long branches leading to a large polytomy.
It could thus be that the estimated inter-clade rate exceeds the intra-clade rate purely because of this conditioning.
This effect might be particularly important early on in the pandemic where branches with as few as two mutations were used to define new clades.
We therefore also included composite clades 19B+, 19B++ and 20A+ containing sequences from 19A and 19B (rooted on 19B), 19A, 19B, 20A, 20B, 20C, 20D (rooted on 19B), and 20A, 20B, 20C, 20D (rooted on 20A).
The estimates for composite clades 19B+ and 20A+ are consistent with the estimates of the individual clades, while apparent rates of 19B++ are considerably higher.
The latter is due to the rapid expansion and subsequent dominance of clade 20A, the first clade with spike mutation $D614G$, and its descendants which rapidly fixed four additional mutations \citep{korber_tracking_2020}.
We argue that this is an early example of an accelerated global rate of evolution due to adaptive evolution, see Discussion.

At the other extreme, some Omicron clades have very low rate estimates which should be interpreted carefully.
Founder genotypes of Omicron clades 21L (BA.2), 22A (BA.4) and 22B (BA.5) are rarely sampled.
These clades showed considerable diversity, including reversions to the reference, soon after their initial discovery \citep{tegally_emergence_2022}.
This diversity and the complex, possibly recombinant, origin of the Omicron clades makes estimating their rates of evolution challenging.

\begin{table*}
\begin{tabular}{l|rrrrrr}
    \hline
    {\bf clade} &  overall rate $[y^{-1}]$ & aa rate $[y^{-1}]$ &  syn rate $[y^{-1}]$ & overall div. &  aa div. &  syn div. \\
    \hline
    19B &     19.46 &    12.37 &      7.09 &        0 &       0 &        0 \\
   19B+ &     18.56 &    11.92 &      6.64 &        0 &       0 &        0 \\
  19B++ &     27.91 &    16.41 &     11.50 &        0 &       0 &        0 \\
  19A &     19.73 &    12.79 &      6.95 &        2 &       1 &        1 \\
  20A &     15.08 &     7.94 &      7.15 &        6 &       3 &        3 \\
    20B &     14.29 &     7.69 &      6.59 &        9 &       5 &        4 \\
    20C &     16.73 &     9.92 &      6.81 &        8 &       5 &        3 \\
   20A+ &     16.30 &     9.40 &      6.90 &        6 &       3 &        3 \\
    20E &     14.04 &     5.65 &      8.39 &       13 &       5 &        8 \\
    20H (Beta) &     15.48 &     9.51 &      5.96 &       21 &      16 &        5 \\
    20I (Alpha) &     11.56 &     5.49 &      6.07 &       30 &      19 &       11 \\
    20J (Gamma) &     12.97 &     6.29 &      6.68 &       35 &      22 &       13 \\
    21D (Eta) &     11.36 &     6.29 &      5.07 &       27 &      13 &       14 \\
    21G (Lambda) &     12.04 &     6.51 &      5.53 &       28 &      19 &        9 \\
    21H (Mu) &     18.25 &    10.05 &      8.20 &       31 &      21 &       10 \\
    21I (Delta) &     16.55 &     8.93 &      7.62 &       30 &      24 &        6 \\
    21J (Delta) &     14.30 &     7.17 &      7.14 &       34 &      28 &        6 \\
    21K (Omicron) &     13.48 &     6.45 &      7.03 &       56 &      46 &       10 \\
    21L (Omicron) &      9.24 &     4.39 &      4.85 &       68 &      50 &       18 \\
    22A (Omicron) &      7.71 &     2.81 &      4.90 &       71 &      52 &       19 \\
    22B (Omicron) &     11.90 &     5.03 &      6.87 &       66 &      50 &       16 \\
           \hline
    \end{tabular}
\caption{{\bf Evolutionary rates estimates from root-to-tip regressions for overall nucleotide changes, amino acid changes, and synonymous changes.}
    The last three columns give the distances of the clade founder sequence from putative MRCA of SARS-CoV-2.
\label{tab:rates}}
\end{table*}

\begin{figure}
    \includegraphics[width=0.5\textwidth]{figures/rate_progression.pdf}
    \caption{{\bf Divergence and evolutionary rates of different Nextstrain clades over time.} Synonymous rates estimates are stable in time and fluctuate around the rate estimates for between clades. Non-synonymous rate estimates are highest for clades 19A to 20C that emerged early in the pandemic.
    \label{fig:rate_progression} }
\end{figure}

@@

 \begin{figure*}[tb]
    \includegraphics[width=\textwidth]{figures/fitness_landscape.pdf}
    \caption{{\bf Landscape of selective constraint along the SARS-CoV-2 genome.}
    Solid lines show sliding window smoothing of estimated mutational tolerance at 1st, 2nd, and 3rd positions for a window size of 20 and 7 (faint lines) sites.
    The markers at the bottom of the panels show qualitatively how many Pango lineages are differ at this position from the Wuhan-Hu-1 reference sequence.
    \label{fig:fitness_landscape}}
 \end{figure*}
